Bayesian Inference from Time Series: Information Loss and the Whittle Likelihood

### 2018-03-14 doc/ms.tex
A^2
More results
PSD of AR(1)
AR(1) writing

@@

We assume that the probability distribution for $\epsilon(t)$ is indeed white
noise. In this case, the probability distribution for $y(t)$ given the
parameters is a gaussian process with mean function of zero and
covariance function given by
\begin{align}
C(\tau) &= A^2\exp\left(-\frac{\omega_0\tau}{2Q}\right)
            \left[\cos(\eta\omega_0\tau)
                + \frac{1}{2\eta Q}\sin(\eta\omega_0\tau)\right]
    \label{eqn:oscillation_covariance}
\end{align}
where $\eta = \sqrt{1 - 1/(4Q^2)}$, $\tau$ is a time separation,
and $Q > 1/2$ (i.e., we assume the oscillation is not overdamped).
A more general expression which allows for $Q \leq 1/2$
is given by \citet{anderson} and \citet{celerite}.
In the limit of large $Q$, the covariance function is approximately
the product of an oscillation and an exponential decay:
\begin{align}
C(\tau) &= A\exp\left(-\frac{\omega_0\tau}{2Q}\right)\cos(\omega_0\tau)
\end{align}
which was the form assumed by \citet{brewer_stello}.
The power spectral density corresponding to
Equation~\ref{eqn:oscillation_covariance} is
\begin{align}
S(\omega) &= \frac{2S_0\omega_0^4}
                  {(\omega^2 - \omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{align}
where $S_0 = A^2/(\omega_0Q)$. This differs from the formula given by
\citet{celerite} by a factor of $\sqrt{2\pi}$ due to different fourier
transform conventions (this formula omits a factor of $1/\sqrt{2\pi}$
in front of the continuous fourier transform).

@@

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{\hspace{2em}}l@{}}
\toprule
$N$         &       Likelihood      &  Mut. Inf. (nats) \\
\hline
100         &       Exact           & $2.488 \pm 0.056$ \\
100         &       Whittle         & $2.007 \pm 0.056$ \\
\bottomrule
\end{tabular}
\caption{The mutual information between the data and the log-period, for
different numbers of datapoints $N$ and for different likelihoods (exact vs.
Whittle).
This is equal to the prior expected reduction in the entropy
in going from the prior to the posterior.\label{tab:oscillation_results}}
\end{table}

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{}}
\toprule
$N$         &       $\Delta$ Mut. Inf. (nats) \\
\hline
100         &       $0.481 \pm 0.057$ \\
\bottomrule
\end{tabular}
\caption{The difference in mutual information between the exact and
Whittle likelihoods. This quantifies the size of the advantage of the exact
likelihood.
The error bars are smaller than what you might expect from
taking differences in Table~\ref{tab:oscillation_results}, due to
the common stream of datasets.\label{tab:oscillation_differences}}
\end{table}

@@

>>>>

\section{Example 1: An AR(1) process}
Consider a continuous AR(1) process, whose covariance function is
\begin{align}
C(\tau) &= \sigma^2 \exp\left(-\frac{|\tau|}{L}\right).
\end{align}
If the process is observed at times $\boldsymbol{t} = \{t_0, ..., t_{N-1}\}$,
the distribution of the discrete distribution for the $\boldsymbol{y}$
is a discrete AR(1) distribution with covariance of the same form.
The joint distribution for $\boldsymbol{y}$ given the parameters
may also be written
\begin{align}
y_i &= \alpha y_{i-1} + \beta n_i
\end{align}
where $\alpha = \exp(-1/L)$, $\beta=\sigma\sqrt{1-\alpha^2}$
is the standard deviation of the
innovations, and the $\{n_i\}$ quantities have independent Normal$(0, 1)$
distributions.

The power spectral density (PSD) of the discrete distribution is
\begin{align}
S(\omega) &= \sigma^2\frac{1 - \alpha^2}{1 + \alpha^2 - 2\alpha\cos(\omega)}.
\end{align}

@@

We assume that the probability distribution for $\epsilon(t)$ is indeed white
noise. In this case, the probability distribution for $y(t)$ given the
parameters is a gaussian process with mean function of zero and
covariance function given by
\begin{align}
C(\tau) &= A^2\exp\left(-\frac{\omega_0\tau}{2Q}\right)
            \left[\cos(\eta\omega_0\tau)
                + \frac{1}{2\eta Q}\sin(\eta\omega_0\tau)\right]
    \label{eqn:oscillation_covariance}
\end{align}
where $\eta = \sqrt{1 - 1/(4Q^2)}$, $\tau$ is a time separation,
and $Q > 1/2$ (i.e., we assume the oscillation is not overdamped).
A more general expression which allows for $Q \leq 1/2$
is given by \citet{anderson} and \citet{celerite}.
In the limit of large $Q$, the covariance function is approximately
the product of an oscillation and an exponential decay:
\begin{align}
C(\tau) &= A^2\exp\left(-\frac{\omega_0\tau}{2Q}\right)\cos(\omega_0\tau)
\end{align}
which was the form assumed by \citet{brewer_stello}.
The power spectral density corresponding to
Equation~\ref{eqn:oscillation_covariance} is
\begin{align}
S(\omega) &= \frac{2S_0\omega_0^4}
                  {(\omega^2 - \omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{align}
where $S_0 = A^2/(\omega_0Q)$. This differs from the formula given by
\citet{celerite} by a factor of $\sqrt{2\pi}$ due to different fourier
transform conventions (this formula omits a factor of $1/\sqrt{2\pi}$
in front of the continuous fourier transform).

@@

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{\hspace{2em}}l@{}}
\toprule
$N$         &       Likelihood      &  Mut. Inf. (nats) \\
\hline
20          &       Exact           & $1.945 \pm 0.057$ \\
50          &       Exact           & $2.263 \pm 0.058$ \\
100         &       Exact           & $2.488 \pm 0.056$ \\
200         &       Exact           & $2.750 \pm 0.054$ \\
\hline
20          &       Whittle         & $1.552 \pm 0.048$ \\
50          &       Whittle         & $1.863 \pm 0.056$ \\
100         &       Whittle         & $2.007 \pm 0.056$ \\
200         &       Whittle         & $2.089 \pm 0.060$ \\
\bottomrule
\end{tabular}
\caption{The mutual information between the data and the log-period, for
different numbers of datapoints $N$ and for different likelihoods (exact vs.
Whittle).
This is equal to the prior expected reduction in the entropy
in going from the prior to the posterior.\label{tab:oscillation_results}}
\end{table}

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{}}
\toprule
$N$         &       $\Delta$ Mut. Inf. (nats) \\
\hline
20          &       $0.393 \pm 0.050$ \\
50          &       $0.400 \pm 0.056$ \\
100         &       $0.481 \pm 0.057$ \\
200         &       $0.661 \pm 0.061$ \\
\bottomrule
\end{tabular}
\caption{The difference in mutual information between the exact and
Whittle likelihoods. This quantifies the size of the advantage of the exact
likelihood.
The error bars are smaller than what you might expect from
taking differences in Table~\ref{tab:oscillation_results}, due to
the common stream of datasets.\label{tab:oscillation_differences}}
\end{table}

@@

\appendix
\section{Notes}

The discrete AR(1) process has covariance function
\begin{align}
C(\tau) &= \sigma^2 \exp\left(-|\tau| / L\right) \\
        &= \sigma^2 \alpha^{|\tau|},
\end{align}
where $\tau$ is the {\em integer} lag.
The power spectral density is therefore
\begin{align}
S(\omega) &= \sum_{\tau=-\infty}^{\infty} C(\tau)\exp(-i \omega \tau) \\
          &= \sum_{\tau=-\infty}^{\infty} \sigma^2 \alpha^{|\tau|}
                                         \exp(-i \omega \tau) \\
          &= \sigma^2 \left(
                \sum_{\tau=-\infty}^{-1} \alpha^{-\tau}\exp(-i \omega \tau) + 
                \sum_{\tau=0}^{\infty} \alpha^{\tau}\exp(-i \omega \tau)
                        \right) \\
          &= \sigma^2 \left(
                \frac{\alpha e^{i\omega}}
                     {1 - \alpha e^{i\omega}}
                 +
                \frac{1}
                     {1 - \alpha e^{-i\omega}}
                        \right) \\
         &= \sigma^2
                \frac{1 - \alpha^2}
                     {1 + \alpha^2 - 2\alpha\cos(\omega)}
\end{align}
